---
title: "Final Project"
author: "Jose Morillo"
date: "Sunday, April 26, 2015"
output: html_document
---
***
### Before starting

My intention with this dataset is identifying variables correlations, 
specially the ones correlated with quality. In order to do so I'll
start by finding the numeric correlations for all variable pairs, and
for each of the significant ones I'll add them to a ggpairs plot and
I'll plot and then refine a histogram. After that I'll continue exploring 
the relation between them in the following sections. The final objective
 will be finding the best predictors for quality.

***
### Data loading and initial exploration

```{r echo=FALSE, message=FALSE, warning=FALSE, packages = FALSE}
options(warn=-1)
wines <- read.csv("wineQualityReds.csv")
str(wines)
wines$X <- NULL
summary(wines)
```

The citric acid amounts found in wine are always below 1 g/dm^3, the density 
variability is very low (0.9901 - 1.0037 g/cm^3), the mean and the median pH 
is 3.31, the alcohol % of volume mean is 10.42 and the median quality is 6.

***
### Find correlations

```{r echo=FALSE, message=FALSE, warning=FALSE, packages = FALSE}
cor(wines)
```
Significant positive and negative correlations found:

fixed.acidity - citric.acid: 0.67170343
fixed.acidity - density: 0.66804729
fixed.acidity - pH: -0.68297819

volatile.acidity - citric.acid: -0.552495685

citric.acid - pH: -0.54190414

free.sulfur.dioxide - total.sulfor.dioxide: 0.667666450

density - alcohol: -0.49617977

quality - alcohol: 0.47616632
quality - volatile.acidity: -0.39055778
quality - citric.acid: 0.22637251
quality - sulphates: 0.25139708

***
### First plots

General exploration and first set of histograms for each variable.

```{r echo=FALSE, message=FALSE, warning=FALSE, packages = FALSE}
library(ggplot2)
library(GGally)
theme_set(theme_minimal(20))
ggpairs(wines, 
        columns = c(1,2,3,6,7,8,9,10,11,12),
        axisLabels = "show", 
        columnLabels = c("f.acid", "vol.acid", "citric.acid", 
                         "f.slf.diox", "t.slf.diox", 
                         "density", "pH", "sulph", 
                         "alcohol", "quality"),
        upper = list(params = c(size = 4)),
        lower = list(params = c(size = 1))) + 
  theme(axis.ticks = element_blank(), 
        axis.text.x = element_blank(), 
        axis.text.y = element_blank())

ggplot(aes(x = fixed.acidity), data = wines) + geom_histogram()
ggplot(aes(x = volatile.acidity), data = wines) + geom_histogram()
ggplot(aes(x = citric.acid), data = wines) + geom_histogram()
ggplot(aes(x = residual.sugar), data = wines) + geom_histogram()
ggplot(aes(x = chlorides), data = wines) + geom_histogram()
ggplot(aes(x = free.sulfur.dioxide), data = wines) + geom_histogram()
ggplot(aes(x = total.sulfur.dioxide), data = wines) + geom_histogram()
ggplot(aes(x = density), data = wines) + geom_histogram()
ggplot(aes(x = pH), data = wines) + geom_histogram()
ggplot(aes(x = sulphates), data = wines) + geom_histogram()
ggplot(aes(x = alcohol), data = wines) + geom_histogram()
ggplot(aes(x = quality), data = wines) + geom_histogram()
```

***
### More refined histograms 

After the previous set of histograms, I'll refine them a little bit more, 
adjusting the limits for x axis using quantiles and xlim and choosing 
a proper binwidth.

```{r echo=FALSE, message=FALSE, warning=FALSE, packages = FALSE}

ggplot(aes(x = fixed.acidity), data = wines) + 
  geom_histogram(binwidth = 1) + 
  xlim(c(quantile(wines$fixed.acidity, 0.01), 
         quantile(wines$fixed.acidity, 0.99)))

ggplot(aes(x = volatile.acidity), data = wines) + 
  geom_histogram(binwidth = 0.1) + 
  xlim(c(quantile(wines$volatile.acidity, 0.01), 
         quantile(wines$volatile.acidity, 0.99)))
  
ggplot(aes(x = citric.acid), data = wines) + 
  geom_histogram(binwidth = 0.1) + 
  xlim(c(quantile(wines$citric.acid, 0.01), 
         quantile(wines$citric.acid, 0.99)))

ggplot(aes(x = residual.sugar), data = wines) + 
  geom_histogram(binwidth = 1) + 
  xlim(c(quantile(wines$residual.sugar, 0.01), 
         quantile(wines$residual.sugar, 0.99)))

ggplot(aes(x = chlorides), data = wines) + 
  geom_histogram(binwidth = 0.05) + 
  xlim(c(quantile(wines$chlorides, 0.01), 
         quantile(wines$chlorides, 0.99)))

ggplot(aes(x = free.sulfur.dioxide), data = wines) + 
  geom_histogram(binwidth = 5) + 
  xlim(c(quantile(wines$free.sulfur.dioxide, 0.01), 
         quantile(wines$free.sulfur.dioxide, 0.99)))

ggplot(aes(x = total.sulfur.dioxide), data = wines) + 
  geom_histogram(binwidth = 20) + 
  xlim(c(quantile(wines$total.sulfur.dioxide, 0.01), 
         quantile(wines$total.sulfur.dioxide, 0.99)))

ggplot(aes(x = density), data = wines) + 
  geom_histogram(binwidth = 0.001) + 
  xlim(c(quantile(wines$density, 0.01), 
         quantile(wines$density, 0.99)))

ggplot(aes(x = pH), data = wines) + 
  geom_histogram(binwidth = 0.1) + 
  xlim(c(quantile(wines$pH, 0.01), 
         quantile(wines$pH, 0.99)))

ggplot(aes(x = sulphates), data = wines) + 
  geom_histogram(binwidth = 0.1) + 
  xlim(c(quantile(wines$sulphates, 0.01), 
         quantile(wines$sulphates, 0.99)))

ggplot(aes(x = alcohol), data = wines) + 
  geom_histogram(binwidth = 0.5) + 
  xlim(c(quantile(wines$alcohol, 0.01), 
         quantile(wines$alcohol, 0.99)))

ggplot(aes(x = quality), data = wines) + 
  geom_histogram(binwidth = 1) + 
  xlim(c(quantile(wines$quality, 0.01), 
         quantile(wines$quality, 0.99)))

summary(wines$quality[wines$chlorides >= 0.25])

```

The vast majority of the wine's salt amount is between 0.05 and 0.1 grams per 
litre. When removing outliers I wondered what the quality for the few ones 
above 0.25 is. After calculating the summary, looks like salt is not a great 
quality predictor. We can also check this with the correlation: -0.128906560.
Density values are very concentrated around 0.997 and the same happens for pH
around 3.4.

***
### Exploring highly correlated variables

I'll now plot scatterplots for the highly correlated pairs of variables that
I found previously, but just for the relations that include at least one 
variable highly correlated to quality. As the correlation is never very high,
I'll add a smoother to see the tendency easily. In the cases in which quality
is not included in the relation, I'll use it to color the dots.

```{r echo=FALSE, message=FALSE, warning=FALSE, packages = FALSE}

wines$qualityFactor <- as.factor(wines$quality)

ggplot(aes(x = fixed.acidity, y = citric.acid, 
           color = qualityFactor, group = 1), data = wines) + 
  geom_point(size = 4, alpha = 1/3) +
  scale_color_brewer(type='div') +
  geom_smooth(method = 'lm') +
  scale_x_continuous(lim = c(quantile(wines$fixed.acidity, 0.01), 
                             quantile(wines$fixed.acidity, 0.99))) +
  scale_y_continuous(lim = c(quantile(wines$wines$citric.acid, 0.01), 
                             quantile(wines$citric.acid, 0.99)), 
                     breaks = seq(0, 0.7, 0.1))

ggplot(aes(x = volatile.acidity, y = citric.acid, 
           color = qualityFactor, group = 1), data = wines) +
  geom_point(size = 4, alpha = 1/3) +
  scale_color_brewer(type='div') +
  geom_smooth(method = 'lm') +
  scale_x_continuous(lim = c(quantile(wines$volatile.acidity, 0.01), 
                             quantile(wines$volatile.acidity, 0.99))) +
  scale_y_continuous(lim = c(quantile(wines$citric.acid, 0.01), 
                             quantile(wines$citric.acid, 0.99)))

ggplot(aes(x = citric.acid, y = pH, 
           color = qualityFactor, group = 1), data = wines) +
  geom_point(size = 4, alpha = 1/3) +
  scale_color_brewer(type='div') +
  geom_smooth(method = 'lm') +
  scale_x_continuous(lim = c(quantile(wines$citric.acid, 0.01), 
                             quantile(wines$citric.acid, 0.99))) +
  scale_y_continuous(lim = c(quantile(wines$pH, 0.01), 
                             quantile(wines$pH, 0.99)))

ggplot(aes(x = density, y = alcohol, 
           color = qualityFactor, group = 1), data = wines) +
  geom_point(size = 4, alpha = 1/3) +
  scale_color_brewer(type='div') +
  geom_smooth(method = 'lm') +
  scale_x_continuous(lim = c(quantile(wines$density, 0.01), 
                             quantile(wines$density, 0.99))) +
  scale_y_continuous(lim = c(quantile(wines$alcohol, 0.01), 
                             quantile(wines$alcohol, 0.99)))

ggplot(aes(x = quality, y = alcohol), data = wines) +
  geom_jitter(size = 4, color = 'red', alpha = 1/3) +
  geom_smooth(method = 'lm') +
  scale_x_continuous(lim = c(quantile(wines$quality, 0.01), 
                             quantile(wines$quality, 0.99))) +
  scale_y_continuous(lim = c(quantile(wines$alcohol, 0.01), 
                             quantile(wines$alcohol, 0.99)))

ggplot(aes(x = quality, y = volatile.acidity), data = wines) +
  geom_jitter(size = 4, color = 'red', alpha = 1/3) +
  geom_smooth(method = 'lm') +
  scale_x_continuous(lim = c(quantile(wines$quality, 0.01), 
                             quantile(wines$quality, 0.99))) +
  scale_y_continuous(lim = c(quantile(wines$volatile.acidity, 0.01), 
                             quantile(wines$volatile.acidity, 0.99)))

ggplot(aes(x = quality, y = citric.acid), data = wines) +
  geom_jitter(size = 4, color = 'red', alpha = 1/3) +
  geom_smooth(method = 'lm') +
  scale_x_continuous(lim = c(quantile(wines$quality, 0.01), 
                             quantile(wines$quality, 0.99))) +
  scale_y_continuous(lim = c(quantile(wines$citric.acid, 0.01), 
                             quantile(wines$citric.acid, 0.99)))

ggplot(aes(x = quality, y = sulphates), data = wines) +
  geom_jitter(size = 4, color = 'red', alpha = 1/3) +
  geom_smooth(method = 'lm') +
  scale_x_continuous(lim = c(quantile(wines$quality, 0.01), 
                             quantile(wines$quality, 0.99))) +
  scale_y_continuous(lim = c(quantile(wines$sulphates, 0.01), 
                             quantile(wines$sulphates, 0.99)))

```

While looking at the fixed.acidity vs citric.acid plot I realised there are 
many wines with citric.acid ~0.5 and some with ~0.25, which I didn't notice 
in the initial histograms. Also, while plotting the scatterplots with the 
quality variable, as it is a discrete variable, I thought about plotting the 
average alcohol/volatile.acidity/citric.acid/sulphates amount per quality 
level, I'll do that in the next section.

```{r echo=FALSE, message=FALSE, warning=FALSE, packages = FALSE}

p1.1 <- ggplot(aes(x = quality, y = alcohol), data = wines) + 
         geom_point(color = 'red', 
                    alpha = 1/2, 
                    position = position_jitter(h = 0)) +
         scale_x_continuous(lim = c(quantile(wines$quality, 0.01), 
                                    quantile(wines$quality, 0.99))) +
         scale_y_continuous(lim = c(quantile(wines$alcohol, 0.01), 
                                    quantile(wines$alcohol, 0.99))) +
         geom_line(stat = 'summary', fun.y = mean)  

p1.2 <- ggplot(aes(x = quality, y = volatile.acidity), data = wines) +
         geom_point(color = 'red', 
                    alpha = 1/2, 
                    position = position_jitter(h = 0)) +
         scale_x_continuous(lim = c(quantile(wines$quality, 0.01), 
                                    quantile(wines$quality, 0.99))) +
         scale_y_continuous(lim = c(quantile(wines$volatile.acidity, 0.01), 
                                    quantile(wines$volatile.acidity, 0.99))) +
         geom_line(stat = 'summary', fun.y = mean)  

p1.3 <- ggplot(aes(x = quality, y = citric.acid), data = wines) +
         geom_point(color = 'red', 
                    alpha = 1/2, 
                    position = position_jitter(h = 0)) +
         scale_x_continuous(lim = c(quantile(wines$quality, 0.01), 
                                    quantile(wines$quality, 0.99))) +
         scale_y_continuous(lim = c(quantile(wines$citric.acid, 0.01), 
                                    quantile(wines$citric.acid, 0.99))) +
         geom_line(stat = 'summary', fun.y = mean)  

p1.4 <- ggplot(aes(x = quality, y = sulphates), data = wines) +
         geom_point(color = 'red', 
                    alpha = 1/2, 
                    position = position_jitter(h = 0)) +
         scale_x_continuous(lim = c(quantile(wines$quality, 0.01), 
                                    quantile(wines$quality, 0.99))) +
         scale_y_continuous(lim = c(quantile(wines$sulphates, 0.01), 
                                    quantile(wines$sulphates, 0.99))) +
         geom_line(stat = 'summary', fun.y = mean)

library(grid)
library(gridExtra)

grid.arrange(p1.1, p1.2, p1.3, p1.4, ncol = 2)

```

Alcohol seems to be the variable that contributes the most to quality. Let's 
see a boxplot of alcohol per quality rank.

Final plots
=====================================================

***
### Boxplot for alcohol and quality

```{r echo=FALSE, message=FALSE, warning=FALSE, packages = FALSE}

ggplot(aes(x = qualityFactor, y = alcohol), data = wines) + 
       geom_boxplot() +
       stat_summary(fun.y = mean, geom = 'point', shape = 4, size = 4) +
       xlab('Quality') +
       ylab('Alcohol (% of volume)') +
       ggtitle('Alcohol vs Quality')
```
```{r}
summary(wines$alcohol[wines$quality == 3])
summary(wines$alcohol[wines$quality == 4])
summary(wines$alcohol[wines$quality == 5])
summary(wines$alcohol[wines$quality == 6])
summary(wines$alcohol[wines$quality == 7])
summary(wines$alcohol[wines$quality == 8])
```

###Description

We can clearly see a trend here. From average wines (quality == 5), the alcohol
content starts to grow in better wines. I added the mean (cross) to the plot, 
this way the representation conveys all the information contained in the
summaries, but much more effectively.

***
### Frequency polygon for quality

```{r echo=FALSE, message=FALSE, warning=FALSE, packages = FALSE}

ggplot(aes(x = quality), data = wines) +
  geom_freqpoly(binwidth = 1) +
  scale_x_continuous(limits = c(3, 8), breaks = seq(3, 8, 1)) +
  xlab('Quality') +
  ylab('Count of wines') +
  ggtitle('Frequency polygon for quality')

summary(wines$quality)

```

###Description

Here we see that most of the wines are labeled as average or good (~6) quality. 
We saw in the variable summary that the median was 6 and the mean was 5.636.
This plot also tells us that this dataset contains limited information on very
good or very bad wines, therefore it would be difficult to extend our 
conclussions with a high degree of certainty to these wines, or create 
accurate predictive models.


***
### Scatterplot for fixed.acidity and density by quality

```{r echo=FALSE, message=FALSE, warning=FALSE, packages = FALSE}

ggplot(aes(x = fixed.acidity, y = density, 
           color = qualityFactor, group = 1), data = wines) + 
       geom_point(size = 3, alpha = 1/2) +
       scale_color_brewer(type='div') +
       geom_smooth(method = 'lm') +
       coord_cartesian(xlim = c(quantile(wines$fixed.acidity, 0.01), 
                                quantile(wines$fixed.acidity, 0.99)), 
                       ylim = c(quantile(wines$density, 0.01), 
                                quantile(wines$density, 0.99))) +
       xlab('Fixed acidity (g / dm^3)') +
       ylab('Density (g / cm^3)') +
       ggtitle('Fixed acidity vs Density')
```

###Description

The fixed acidity increases with the density, but also, specially after adding
the smoother, we can see than the highest quality wines (bluish) are usually
below this line indicating lower density for these wines. I chose this plot
to see the relation between two of the variables with a higher correlation
colored by quality, and the color helped to discover that density correlates
negatively with wine quality. We can visually see this with the following
extra boxplot.


***
### Extra: Density vs Quality boxplot

```{r echo=FALSE, message=FALSE, warning=FALSE, packages = FALSE}

ggplot(aes(x = qualityFactor, y = density), data = wines) + 
       geom_boxplot() +
       stat_summary(fun.y = mean, geom = 'point', shape = 4, size = 4) +
       xlab('Quality') +
       ylab('Density (g / cm^3)') +
       ggtitle('Density vs Quality')

```

Reflection
=====================================================

After exploring the dataset I can say, looking at the most correlated 
variables, that the most density a wine has the less alcohol it has, being
this last variable the single highest predictor for wine quality, although 
still not being very highly correlated. Also, most of the wines were rated 
slightly above medium quality (5). I'd say that with these variables alone 
it would be difficult to generate a highly reliable model to predict quality. 
Adding smoothers to the plots was a great improvement as it wasn't easy to 
see the trend in most of them. Looking for similar datasets I found some 
other variables that could have been useful predictors in our case, such as 
type of grapes or amount of rain for that year, along with some other weather 
and soil conditions. Adding those variables to the dataset could contribute to
better models.